---
title: "Chapter 2 Worksheet"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = FALSE)
```

# Twin Study
Read case study 2.1.2. Read in the data.

```{r, echo=TRUE}
library(Sleuth3)
data("case0202")
```

1. Create a vector containing the difference in the left-hippocampus in `Affected` and `Unaffected` twins.

```{r}
diff <- case0202$Unaffected - case0202$Affected
```


2. Use the function `t.test()` to run a one-sample $t$-test against a null hypothesis of a mean difference of 0. You should read about this function with `help(t.test)`.

```{r}
t.test(diff)
```

3. Now let's do this $t$-test the hard way. What are the mean and standard deviations of the difference in the left hippocampus between unaffected and affected twins? What is the sample size?

```{r}
## Mean
mean(diff)

## Standard Deviation
sd(diff)

## Sample Size
length(diff)
```

4. What is the $t$-statistic? Save it as the variable `tstat`.

```{r}
tstat <- mean(diff) / (sd(diff) / sqrt(length(diff)))
tstat
```

5. What is the degrees of freedom? Save it as a variable `degfr`.

```{r}
degfr <- length(diff) - 1
degfr
```

6. Use the `pt()` function to get the same $p$-value as in part 2.

```{r}
2 * pt(q = -abs(tstat), df = degfr)
```

# Beak Study

Read about Case 2.1.1. Read it into R

```{r, echo=TRUE}
library(Sleuth3)
data(case0201)
case0201$Year <- factor(case0201$Year)
```

1. Make a boxplot of `Year` (x-axis) against `Depth` (y-axis). Do the Depths from the two years look like they might have the same variance?

```{r}
## Yes
library(ggplot2)
qplot(x = Year, y = Depth, data = case0201, geom = "boxplot")
```

2. Now consider the following output of `t.test()`.

```{r, echo=TRUE}
t.test(formula = Depth ~ Year, data = case0201, var.equal = TRUE)
```

Write out a statistical conclusion from the above procedure.

```{block}
We reject the null hypothesis of same mean beak size in the two years ($p$ < 0.001) and conclude that beak size is different. Estimated difference in means is 0.67 (95% CI of 0.38 to 0.96).
```


