---
title: "Multi-factorial ANOVA"
author: "David Gerard"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: "metropolis"
    slide_level: 2
urlcolor: "blue"
---

```{r setup, include=FALSE}
ggplot2::theme_set(
  ggplot2::theme_bw()
  )
knitr::opts_chunk$set(echo = FALSE,
                      fig.height = 2.5, 
                      fig.width = 4, 
                      fig.align = "center")
```

## Objectives

- Brief summary of how we generalize to more than two factors.

## Poker Case Study

- `Skill`: `Expert`, `Average`
- `Hand`: `Bad`, `Neutral`, `Good`
- `Limit`: `Fixed`, `None`
- `Cash`: Final cash balance

Source: G. Meyer, M. von Meduna, T. Brosowski, T. Hayer (2012). "Is poker a Game of Skill or Chance? A Quasi-Experimental Study," Journal of Gambling Studies, Online First DOI 10.1007/s10899-012-9327-8

## Data

```{r, echo = TRUE}
load("../data/poker_skill.rdata")
head(poker)
```



```{r, echo = FALSE}
library(ggplot2)
library(ggthemes)
# library(tidyverse)
# poker <- read.table("../data/poker_skill.dat")
# 
# poker %>%
#   transmute(Skill = factor(recode(V1, "1" = "Expert", "2" = "Average")),
#             Hand  = factor(recode(V2, "1" = "Bad", "2" = "Neutral", "3" = "Good")),
#             Limit = factor(recode(V3, "1" = "Fixed", "2" = "None")),
#             Cash = V4) %>%
#   as_tibble() ->
#   poker
# 
# save(poker, file = "../data/poker_skill.rdata")
```

## Full Interaction Model

- $Y_{ijk\ell} = \mu + \alpha_i + \beta_j + \gamma_k + (\alpha\beta)_{ij} + (\alpha\gamma)_{ik} + (\beta\gamma)_{jk} + (\alpha\beta\gamma)_{ijk} + \epsilon_{ijk\ell}$

- $\mu$: baseline mean.

- $\alpha_i$, $\beta_j$, $\gamma_k$: main effects

- $(\alpha\beta)_{ij}$, $(\alpha\gamma)_{ik}$, $(\beta\gamma)_{jk}$: two-way interactions.

- $(\alpha\beta\gamma)_{ijk}$: three-way interaction

## Interaction Plots

- One variable on $x$-axis

- One variable as grouping variable.

- One variable distinguish facets

## Interaction Plots

- Two-way interaction without three-way interaction: the effects change, but the interaction effect is the same between any pair of factors as the level of the third factor changes.

- Three way interaction: the interaction effect changes as we plot at different levels of the third factors.

## No Interactions

- Facets are $z$

```{r}
x <- c(1, 1, 1, 1, 0, 0, 0, 0)
y <- c(1, 1, 0, 0, 1, 1, 0, 0)
z <- c(1, 0, 1, 0, 1, 0, 1, 0)
w <- x + y + z
meandf <- data.frame(x = as.factor(x), y = as.factor(y), z = as.factor(z), w = w)

ggplot(data = meandf, mapping = aes(x = x, y = w, color = y, group = y)) +
  geom_point() +
  geom_line() +
  facet_grid(.~z) +
  scale_color_colorblind()
```

## Two-way interactions ($x$ and $y$ only), no three-way interactions

- Facets are $z$

```{r}
x <- c(1, 1, 1, 1, 0, 0, 0, 0)
y <- c(1, 1, 0, 0, 1, 1, 0, 0)
z <- c(1, 0, 1, 0, 1, 0, 1, 0)
w <- x + y + z + x*y
meandf <- data.frame(x = as.factor(x), y = as.factor(y), z = as.factor(z), w = w)

ggplot(data = meandf, mapping = aes(x = x, y = w, color = y, group = y)) +
  geom_point() +
  geom_line() +
  facet_grid(.~z) +
  scale_color_colorblind()
```

## Two-way interactions ($x$ and $y$ only), no three-way interactions

- Facets are $y$

```{r}
ggplot(data = meandf, mapping = aes(x = x, y = w, color = z, group = z)) +
  geom_point() +
  geom_line() +
  facet_grid(.~y) +
  scale_color_colorblind()
```

## Two-way interactions ($x$ and $y$ only), no three-way interactions

- Facets are $y$

```{r}
ggplot(data = meandf, mapping = aes(x = z, y = w, color = x, group = x)) +
  geom_point() +
  geom_line() +
  facet_grid(.~y) +
  scale_color_colorblind()
```

## Two-way interactions ($x$ and $y$ only), no three-way interactions

- Facets are $x$

```{r}
ggplot(data = meandf, mapping = aes(x = z, y = w, color = y, group = y)) +
  geom_point() +
  geom_line() +
  facet_grid(.~x) +
  scale_color_colorblind()
```


## Two way interactions (between all variables), no three way

- Facets are $z$

```{r}
x <- c(1, 1, 1, 1, 0, 0, 0, 0)
y <- c(1, 1, 0, 0, 1, 1, 0, 0)
z <- c(1, 0, 1, 0, 1, 0, 1, 0)
w <- x + y + z + x*y + x*z + y*z 
meandf <- data.frame(x = as.factor(x), y = as.factor(y), z = as.factor(z), w = w)

ggplot(data = meandf, mapping = aes(x = x, y = w, color = y, group = y)) +
  geom_point() +
  geom_line() +
  facet_grid(.~z) +
  scale_y_continuous(breaks = 1:6) +
  scale_color_colorblind()
```



## Three-way interactions

- Facets are $z$

```{r}
x <- c(1, 1, 1, 1, 0, 0, 0, 0)
y <- c(1, 1, 0, 0, 1, 1, 0, 0)
z <- c(1, 0, 1, 0, 1, 0, 1, 0)
w <- x + y + z + x*y + x*z + y*z + x * y * z
meandf <- data.frame(x = as.factor(x), y = as.factor(y), z = as.factor(z), w = w)

ggplot(data = meandf, mapping = aes(x = x, y = w, color = y, group = y)) +
  geom_point() +
  geom_line() +
  facet_grid(.~z) +
  scale_y_continuous(breaks = 1:7) +
  scale_color_colorblind()
```


## Interaction Plots
\small
```{r, echo = TRUE, fig.height=2.5}
qplot(x = Skill, y = Cash, color = Hand, 
      group = Hand, facets = . ~ Limit, 
      geom = "blank", data = poker) +
  stat_summary(fun.y = mean, geom = "line")
```
\normalsize

## Interaction Plots
\small
```{r, echo = TRUE, fig.height=2.5}
qplot(x = Hand, y = Cash, color = Skill, 
      group = Skill, facets = . ~ Limit, 
      geom = "blank", data = poker) +
  stat_summary(fun.y = mean, geom = "line")
```
\normalsize

## Interaction Plots
\small
```{r, echo = TRUE, fig.height=2.5}
qplot(x = Limit, y = Cash, color = Skill, 
      group = Skill, facets = . ~ Hand, 
      geom = "blank", data = poker) +
  stat_summary(fun.y = mean, geom = "line")
```
\normalsize

## Interaction Plots
\small
```{r, echo = TRUE, fig.height=2.5}
qplot(x = Skill, y = Cash, color = Limit, 
      group = Limit, facets = . ~ Hand, 
      geom = "blank", data = poker) +
  stat_summary(fun.y = mean, geom = "line")
```
\normalsize

## Interaction Plots
\small
```{r, echo = TRUE, fig.height=2.5}
qplot(x = Hand, y = Cash, color = Limit, 
      group = Limit, facets = . ~ Skill, 
      geom = "blank", data = poker) +
  stat_summary(fun.y = mean, geom = "line")
```
\normalsize

## Interaction Plots
\small
```{r, echo = TRUE, fig.height=2.5}
qplot(x = Limit, y = Cash, color = Hand, 
      group = Hand, facets = . ~ Skill, 
      geom = "blank", data = poker) +
  stat_summary(fun.y = mean, geom = "line")
```
\normalsize

## In R
```{r, echo = TRUE}
## No evidence of a three-way interaction
aout_cell <- aov(Cash ~ Skill * Hand * Limit, data = poker)
summary(aout_cell)
```

## In R
```{r, echo = TRUE}
aout_sub <- aov(Cash ~ Skill + Hand + Limit + 
                  Skill * Hand + Skill * Limit,
                data = poker)
summary(aout_sub)
```

## Residual Plots

```{r, echo = TRUE, fig.height=2.5}
fvec <- fitted(aout_sub)
rvec <- resid(aout_sub)
qplot(fvec, rvec, xlab = "Fitted", ylab = "Residuals")
```


## Still interesting 
\footnotesize
```{r, echo = TRUE}
t(t(coef(aout_sub)))
```
\normalsize

## Interpretations
\begin{center}
\footnotesize
\begin{tabular}{l|l}
Variable & Estimate\\
\hline
$\mu$       & 5.400\\
$\alpha_1$  & 0 (by definition)\\
$\alpha_2$  & 1.404\\
$\beta_1$   & 0 (by definition)\\
$\beta_2$   & 5.180\\
$\beta_3$   & 9.270\\
$\gamma_1$  & 0 (by definition)\\
$\gamma_2$  & -1.910\\
$(\alpha\beta)_{11}$ & 0 (by definition)\\
$(\alpha\beta)_{12}$ & 0 (by definition)\\
$(\alpha\beta)_{13}$ & 0 (by definition)\\
$(\alpha\beta)_{21}$ & 0 (by definition)\\
$(\alpha\beta)_{22}$ & -1.440\\
$(\alpha\beta)_{23}$ & -4.124\\
$(\alpha\gamma)_{11}$ & 0 (by definition)\\
$(\alpha\gamma)_{12}$ & 0 (by definition)\\
$(\alpha\gamma)_{21}$ & 0 (by definition)\\
$(\alpha\gamma)_{22}$ & 2.520\\
\end{tabular}
\end{center}

## Interpretations
- $\alpha$ for skill level
- $\beta$ for hand
- $\gamma$ for limit

- Mean difference between experts with good hands and experts with bad hands is
\begin{align*}
\mu& + \alpha_2 + \beta_3 + \gamma_k + (\alpha\beta)_{23} + (\alpha\gamma)_{2k}\\
&-\left[\mu + \alpha_2 + \beta_1 + \gamma_k + (\alpha\beta)_{21} + (\alpha\gamma)_{2k}\right]\\
&= \mu + \alpha_2 + \beta_3 + \gamma_k + (\alpha\beta)_{23} + (\alpha\gamma)_{2k}\\ &-\left[\mu + \alpha_2 + 0 + \gamma_k + 0 + (\alpha\gamma)_{2k}\right]\\
&= \beta_3 + (\alpha\beta)_{23}\\
&= 9.270 - 4.124 = 5.146
\end{align*}


## Interpretations
- $\alpha$ for skill level
- $\beta$ for hand
- $\gamma$ for limit

- Mean difference between non-experts with good hands and non-experts with bad hands is
\begin{align*}
\mu& + \alpha_1 + \beta_3 + \gamma_k + (\alpha\beta)_{13} + (\alpha\gamma)_{1k}\\
&-\left[\mu + \alpha_1 + \beta_1 + \gamma_k + (\alpha\beta)_{11} + (\alpha\gamma)_{1k}\right]\\
&= \beta_3\\
&= 9.270
\end{align*}

- Experts benefit less from a quality hand?

## Interpretations

- Similarly ideas for limit.

- SkillExpert:LimitNone > 0 indicates that experts benefit **more** from not having a limit.




