---
title: "Two-way ANOVA Main Effects"
author: "David Gerard"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: "metropolis"
    slide_level: 2
urlcolor: "blue"
---

```{r setup, include=FALSE}
ggplot2::theme_set(
  ggplot2::theme_bw()
  )
knitr::opts_chunk$set(echo = FALSE,
                      fig.height = 2.5, 
                      fig.width = 4, 
                      fig.align = "center")
```

## Objectives

- Testing for Main Effects
- Chapter 13 in the book.

## Pygmalion Effect Case Study

- Pygmalion Effect: high expectations of a supervisor translate into improved performance of subordinate.
- A company of soldiers contains three platoons each.
- Within each company, one platoon was randomly selected to be the "Pygmalion platoon."
- The platoon leader in the Pygmalion platoon was told by the army psychologist that his platoon was predicted to be superior.
- At end of basic training, all soldiers in each platoon were given a skill test.
- Data consist of average scores for each platoon.

## The data
```{r, echo = TRUE}
library(Sleuth3)
data("case1302")
head(case1302)
```

## Two-way ANOVA model: The additive model

- Model: $Y_{ijk} = \mu + \alpha_i + \beta_j + \epsilon_{ijk}$ \bigskip
- $Y_{ijk}$: Value of observational unit $j$ of group $i$ of the first categorical variable and group $j$ of the second categorical variable. \bigskip\pause
- $\mu$: baseline value.\bigskip
- $\alpha_i$: Additive effect of being in group $i$ in categorical variable 1.\bigskip
- $\beta_j$: Additive effect of being in group $j$ in categorical variable 2.\bigskip\pause
- $\epsilon_{ijk}$: Individual-specific noise for observational unit $k$ of group $i$ of the first categorical variable and group $j$ of the second categorical variable. Assumed to have mean 0 and variance $\sigma^2$.
- $\sigma^2$ is assumed to be the **same** for all observational units of all groups

## Main Effects

- The $\alpha_i$'s are called the "main effects" for company (as opposed to interaction effects). 

- The $\beta_j$'s are called the "main effects" for treatment.

- We are not directly interested in the $\alpha_i$'s. The reasearchers just thought that there would be a lot of variability between companies and so wanted to control for this.

- We want to know if the treatment is effective.

- $H_0: \beta_i = 0$ for all $i$.

- $H_A: \beta_i \neq 0$ for some $i$.

## Our Specific Case

- $J = 2$ since there are two groups (Pygmalion and Control)

- $\beta_1 = 0$ by assumption.

- $\beta_2$ is the treatment effect.

- $H_0: \beta_2 = 0$

- $H_A: \beta_2 \neq 0$

## Reduced (Company Only) Model

- Mean estimates

```{r}
suppressMessages(library(tidyverse))
library(ggthemes)
aout_com <- aov(Score ~ Company, data = case1302)

case1302$est_com <- aout_com$fitted.values
case1302$lower <- as.numeric(case1302$Company) - 0.2
case1302$upper <- as.numeric(case1302$Company) + 0.2

ggplot(data = case1302, mapping = aes(x = Company, y = Score, color = Treat)) +
  geom_point() +
  ggthemes::scale_color_colorblind() +
  geom_segment(mapping = aes(x = lower, xend = upper, y = est_com, yend = est_com), color = "blue")
```

## Reduced (Company Only) Model

- Find the Residuals

```{r}
ggplot(data = case1302, mapping = aes(x = Company, y = Score, color = Treat)) +
  geom_point() +
  ggthemes::scale_color_colorblind() +
  geom_segment(mapping = aes(x = lower, xend = upper, y = est_com, yend = est_com)) +
  geom_segment(aes(x = Company, xend = Company, y = Score, yend = est_com), col = "blue", lty = 1, alpha = 1/2, lwd = 1)
```

## Reduced (Company Only) Model

- Find the Residuals (jittered for easier viewing)

```{r}
case1302$com_jit <- jitter(as.numeric(case1302$Company))
ggplot(data = case1302, mapping = aes(x = com_jit, y = Score, color = Treat)) +
  geom_point() +
  ggthemes::scale_color_colorblind() +
  geom_segment(mapping = aes(x = lower, xend = upper, y = est_com, yend = est_com)) +
  geom_segment(aes(x = com_jit, xend = com_jit, y = Score, yend = est_com), col = "blue", lty = 1, alpha = 1/2, lwd = 1) +
  xlab("Company") +
  scale_x_continuous(breaks = 1:10, labels = levels(case1302$Company))
```


## Full (Additive) Model

- Mean Estimates

```{r}
aout_add <- aov(Score ~ Company + Treat, data = case1302)

case1302$est_add <- aout_add$fitted.values

ggplot(data = case1302, mapping = aes(x = Company, y = Score, color = Treat)) +
  geom_point() +
  ggthemes::scale_color_colorblind() +
  geom_segment(mapping = aes(x = lower, xend = upper, y = est_add, yend = est_add))
```


## Full (Additive) Model

- Find the Residuals

```{r}
ggplot(data = case1302, mapping = aes(x = Company, y = Score, color = Treat)) +
  geom_point() +
  ggthemes::scale_color_colorblind() +
  geom_segment(mapping = aes(x = lower, xend = upper, y = est_add, yend = est_add)) +
  geom_segment(aes(x = Company, xend = Company, y = Score, yend = est_add), col = "blue", lty = 1, alpha = 1/2, lwd = 1)
```

## Full (Additive) Model

- Find the Residuals (jittered for easier viewing)

```{r}
ggplot(data = case1302, mapping = aes(x = com_jit, y = Score, color = Treat)) +
  geom_point() +
  ggthemes::scale_color_colorblind() +
  geom_segment(mapping = aes(x = lower, xend = upper, y = est_add, yend = est_add)) +
  geom_segment(aes(x = com_jit, xend = com_jit, y = Score, yend = est_add), col = "blue", lty = 1, alpha = 1/2, lwd = 1) +
  xlab("Company") +
  scale_x_continuous(breaks = 1:10, labels = levels(case1302$Company))
```

## Sum of Squared Residuals

```{r}
n <- nrow(case1302)
Ival <- nlevels(case1302$Company)
Jval <- nlevels(case1302$Treat)
rss_full <- sum(aout_add$residuals^2)
df_full <- n - Ival - Jval + 1
rss_reduced <- sum(aout_com$residuals^2)
df_reduced <- n - Ival
ess <- rss_reduced - rss_full
df_extra <- df_reduced - df_full
fstat <- (ess / df_extra) / (rss_full / df_full)
```

- $RSS_{full}$ = `r rss_full`

- $df_{full}$ = $n - I - J + 1$ = `r n` - `r Ival` -`r Jval` + 1 = `r df_full` \pause

- $RSS_{reduced}$ = `r rss_reduced`

- $df_{reduced}$ = $n - I$ = `r n` - `r Ival` = `r df_reduced` \pause

- $ESS = RSS_{reduced} - RSS_{reduced}$ = `r ess`

- $df_{extra} = df_{reduced} - df_{full}$ = `r df_extra`\pause

- $F$-statistic = $\frac{ESS / df_{extra}}{RSS_{full} / df_{full}}$ = `r fstat`.

## Compare observed `r fstat` to a $F_{1, 18}$ Distribution

```{r}
x <- seq(0, 9, length = 100)
y <- df(x, 1, 18)

which_x <- x > fstat
polydf <- data.frame(polyx = c(min(x[which_x]), x[which_x], max(x[which_x]), min(x[which_x])),
                     polyy = c(0, y[which_x], 0, 0))
qplot(x, y, geom = "line") +
  geom_polygon(data = polydf, mapping = aes(x = polyx, y = polyy), 
               fill = "blue", alpha = 3/4, color = "blue", lwd = 1)
```


## Compare observed `r fstat` to a $F_{1, 18}$ Distribution

```{r, echo = TRUE}
pf(7.84, df1 = 1, df2 = 18, lower.tail = FALSE)
```

- Moderate evidence to suggest a treatment effect.

## In R

```{r, echo = TRUE}
aout_add <- aov(Score ~ Company + Treat, data = case1302)
anova(aout_add)
```

## Get out Coefficient Estimates

```{r, echo = TRUE}
t(t(coef(aout_add)))
```

## Get Confidence Intervals of Effects

```{r, echo = TRUE}
confint(aout_add)
```




