---
title: "Data"
author: "David Gerard"
date: "`r Sys.Date()`"
bibliography: "./bib/614_bib.bib"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Most of the datasets I used come from the `Sleuth3` package, which you can install and load into R using:
```{r, eval = FALSE}
install.packages("Sleuth3")
library(Sleuth3)
```

Many other datasets come from the wonderfully curated list made by Larry Winner at the University of Florida: http://users.stat.ufl.edu/~winner/datasets.html. I'll mark these with a "(W)" by the name of the file.


I will limit this list to those datasets that aren't in the `Sleuth3` package.

- [BIRTHWEIGHT.csv](./data/BIRTHWEIGHT.csv): This file comes from @rosner2015fundamentals. It contains the birthweights of 1000 infants from Boston City Hospital. The variables in this dataset are
    - `index`: The order number of the infant.
    - `weight`: The weight of the infant (in ounces).
    
- [espresso2.rdata](./data/espresso2.rdata) (W): The results of the experiment in @masella2015new to measure the effect of temperature and pressure on foam index in a new method to make espresso. These data were simulated from summary statistics. Variables include
    - `foamIndx`:  The ratio between the foam and liquid volume (expressed as a percent) measured 30s after extraction.
    - `trt_id`: A label for each treatment group.
    - `tempC`: The temperature of the water in Celsius (75, 85, and 90).
    - `prssBar`: The pressure in bars (15 and 20).

- [lsd.rdata](./data/lsd.rdata) (W): In @wagner1968correlation, the authors intravenously injected d-lysergic acid diethylamide (better known as LSD) into seven volunteer male subjects. They then instructed the subjects to solve a set of simple arithmetic problems as quickly as possible during a three minute period. Variables include:
    - `conc`: Tissue concentration of LSD in nanograms of LSD per mL.
    - `score`: Performance score on the arithmetic questions as a percent of a control score.
    
- [poker_skill.rdata](./data/poker_skill.rdata) (W): @meyer2013poker investigated the effects of skill, hand quality, and bidding limits on winnings. These data were simulated from summary statistics. Variables include:
    - `Skill`: Either `Expert` or `Average`.
    - `Hand`: Either `Bad`, `Neutral`, or `Good`.
    - `Limit`: Either `Fixed` or `None`.
    - `Cash`: The final cash balance. 
    
- [sardine.rdata](./data/sardine.rdata) (W): The `sardine` data frame from @clarke1936variations contains the number of vertebrae and the location of 12,858 sardines. Variables include:
    - `location`: The location of the sardine. `1` = Alaska, `2` = British Columbia, `3` = San Francisco, `4` = Monterey, `5` = San Pedro, and `6` = San Diego.
    - `number`: The number of vertebrae of the sardine.
    
- [trek.csv](./data/trek.csv): The `trek` data frame consists of word counts and proportion of words spoken by the main characters from the excellent television series [Star Trek: The Next Generation](https://en.wikipedia.org/wiki/Star_Trek:_The_Next_Generation). I obtained the original dataset [here](https://github.com/RMHogervorst/TNG). The variables in this dataset are
    - `season` The season of the show.
    - `character` The character of the show.
    - `count` The raw word-counts for a given character in a given season.
    - `prop` The proportion of words spoken by a given character in a given season.
    - `logprop` The log of `prop`.


# References
