---
title: "02 Worksheet"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

Read in the data from Case Study 1.1.2 from the `Sleuth3` R package.

```{r}
library(Sleuth3)
data("case0102")
head(case0102)
```

1. Create a boxplot for `Salary` vs `Sex`

```{r, echo=FALSE}
library(ggplot2)
qplot(x = Sex, y = Salary, data = case0102, geom = "boxplot")
```

2. Calculate the mean salary for just the females.

```{r, echo = FALSE}
femave <- mean(case0102$Salary[case0102$Sex == "Female"])
femave
```


3. Calculate the mean salary for just the males.

```{r, echo = FALSE}
malave <- mean(case0102$Salary[case0102$Sex == "Male"])
malave
```

4. What is the difference in average salaries between males and females?

```{r, echo = FALSE}
malave - femave
```


5. Now we will walk through simulating the randomization distribution.

a. The function `sample()` will permute the the elements of a vector. Try it out on `case0102$Sex`
```{r}
sample(case0102$Sex)
```

b. Take one permutation of `Sex` and recalculate the average salary difference between males and females.

c. To automate this, consider the following code using the `replicate()` function
```{r}
sample_vec <- replicate(n = 100,
                        expr = {
                          a <- rnorm(1)
                          b <- rnorm(2)
                          diff <- b - a
                          diff
                        }
                        )
head(sample_vec)
```
Whatever is placed within the braces "{}" will be run `n` times. Anything printed (in this case `diff`) will be returned and placed in `sample_vec`. Now modify this code to create a permutation distribution.

```{r, echo=FALSE}
sample_vec <- replicate(n = 1000,
                        expr = {
                          newsex <- sample(case0102$Sex)
                          a <- mean(case0102$Salary[newsex == "Female"])
                          b <- mean(case0102$Salary[newsex == "Male"])
                          diff <- b - a
                          diff
                        }
                        )
```

d. Plot the permutation distribution and compare to the observed value calculated in 4.

```{r, echo = FALSE}
qplot(sample_vec, geom = "histogram", bins = 30) +
  geom_vline(xintercept = malave - femave, col = 2, lty = 2)
```





