---
title: "Multiple Regression"
author: "David Gerard"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    slide_level: 2
    theme: "metropolis"
urlcolor: "blue"
---

```{r setup, include=FALSE}
ggplot2::theme_set(
  ggplot2::theme_bw()
  )
knitr::opts_chunk$set(echo = FALSE,
                      fig.height = 2.3, 
                      fig.width = 3.7, 
                      fig.align = "center")
```


```{r}
set.seed(1)
library(ggplot2)
```

## Objectives

- Introduce Multiple Linear Regression

- Chapters 9 and 10 in the book.

## Brain Size

- What variables are associated with brain weight?

- Collected information on 96 different species.

- We know that body weight is already associated with brain weight, 
    
    - So what variables are associated with brain weight **after controlling for body weight**.

- Possible variables: Body weight (kg), gestation period (days), litter size

## Brain Size

```{r, echo = TRUE}
library(Sleuth3)
data("case0902")
head(case0902)
```

## Simple Linear Regression

- One quantitative response variable ($Y$).

- One quantitative explanatory variable ($X$).

- The mean of $Y$ is a linear function of $X$.

- Model the conditional distribution of $Y$ given $X$.

- $Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$

## Simple Linear Regression

```{r}
qplot(log(Body), log(Brain), data = case0902)
```

## Simple Linear Regression

```{r}
qplot(log(Body), log(Brain), data = case0902) +
  geom_smooth(method = "lm", se = FALSE)
```

## Multiple Linear Regression Model

- One quantitative response ($Y$).

- **Multiple** quantitative explanatory variables ($X_1, X_2,\ldots,X_p$).

- $Y_i = \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} +\cdots+\beta_pX_{pi} + \epsilon_i$\pause

- E.g. $X_{2i}$ is the value of the second explanatory variable for observational unit $i$.

- E.g., when we have two explanatory variables, this equation is 
$$
Y_i = \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + \epsilon_i
$$

- $\epsilon_i$ is still ideally normally distributed with mean 0 and constant variance $\sigma^2$.



## Multiple Linear Regression: Interpreting Coefficients

- Consider Model $Y_i = \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + \epsilon_i$

- To interpret $\beta_1$ in a **randomized experiment**:

    - Conceptually fix $X_{2i}$ at a value.
    
    - Add one to $X_{1i}$
    
    - $Y_i$ changes by $\beta_1$
    
- $\beta_1$ is how much $Y_i$ increases when we add one to $X_{1i}$ but keep $X_{2i}$ fixed.

- "A one-unit increase in light intensity causes the mean number of flowers to increase by $\beta_1$."


## Multiple Linear Regression: Interpreting Coefficients

- Consider Model $Y_i = \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + \epsilon_i$

- To interpret $\beta_1$ in an **observational study**:

    - The $X_i$'s cannot be fixed independently of one another.
    
    - Consider a subpopulation that has the same values of the $X_{j}$'s, where $j\neq 1$. Then the expected difference in means between species that differ in $X_1$ only by one is $\beta_1$.
    
- $\beta_1$ is the expected difference in $Y$'s when we compare species with $X_{1}$ and $X_{1} + 1$.
    
- "For any subpopulation of mammal species with the same body weight, species with a one-day longer gestation length tend to have a mean brain-weight $\beta_1$ larger."

## A useful notation

- $\mu(brain | gestation) = \beta_0 + \beta_1 gestation$

    - The mean of brain is equal to $\beta_0$ plust $\beta_1$ times the gestation time.

- $\mu(brain|gestation, body) = \beta_0 + \beta_1 gestation + \beta_2 body$

    - The mean of brain is equal to $\beta_0$ plust $\beta_1$ times the gestation time plus $\beta_2$ times the body weight.
    
## Interpretation of coefficients changes when the model differs.

- $\mu(brain | gestation) = \beta_0 + \beta_1 gestation$

    - $\beta_1$ is the mean difference in brain weight as we compare different gestation periords 1 day apart **in the population of all mammal species**.
    
- $\mu(brain|gestation, body) = \beta_0 + \beta_1 gestation + \beta_2 body$

    - $\beta_1$ is the mean difference in brain weight as we compare different gestation periords 1 day apart **in subpopulations that have the same body weight**.
    
## Interpretation of $\beta_1$ **without** $X_2$ in model

- $\mu(brain | gestation) = \beta_0 + \beta_1 gestation$

- Slope looks negative

```{r}
library(ggthemes)
beta0 <- 2
beta2 <- -3
beta1 <- 1
x2 <- rep(1:10, 5)
x1 <- rep(0:9, 5) + rep(1:5, each = 10)
y  <- beta0 + beta1 * x1 + beta2 * x2
qplot(x1, y) + geom_smooth(method = "lm", se = FALSE) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank())
```

## Interpretation of $\beta_1$ **with** $X_2$ in model

- $\mu(brain | gestation, body) = \beta_0 + \beta_1 gestation + \beta_2 body$

- Slope looks positive at each level of $X_2$

```{r}
library(ggthemes)
beta0 <- 2
beta2 <- -3
beta1 <- 1
x2 <- rep(1:10, 5)
x1 <- rep(0:9, 5) + rep(1:5, each = 10)
y  <- beta0 + beta1 * x1 + beta2 * x2
x2 <- as.factor(x2)
qplot(x1, y, color = x2) + 
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  guides(color = guide_legend(
    keywidth=0.1,
    keyheight=0.1,
    default.unit="inch")
    )
```

## Interpretation of $\beta_1$ **with** $X_2$ in model

- $\mu(brain | gestation) = \beta_0 + \beta_1 gestation + \beta_2 body$

- Slope looks positive at each level of $X_2$

```{r}
library(ggthemes)
beta0 <- 2
beta2 <- -3
beta1 <- 1
x2 <- rep(1:10, 5)
x1 <- rep(0:9, 5) + rep(1:5, each = 10)
y  <- beta0 + beta1 * x1 + beta2 * x2
x2 <- as.factor(x2)
qplot(x1, y, color = x2) + 
  geom_smooth(method = "lm", se = FALSE) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  guides(color = guide_legend(
    keywidth=0.1,
    keyheight=0.1,
    default.unit="inch")
  )



# beta0 <- 2
# beta2 <- -3
# beta1 <- 1
# dfdat <- expand.grid(0:14, 0:10)
# x2 <- dfdat[, 2]
# x1 <- dfdat[, 1]
# y  <- beta0 + beta1 * x1 + beta2 * x2
# x2 <- as.factor(x2)
# 
# pl <- qplot(x1, y, color = x2, geom = "blank") +
#   geom_smooth(method = "lm", se = FALSE) +
#   theme(axis.text = element_blank(),
#         axis.ticks = element_blank()) +
#   guides(color = guide_legend(
#     keywidth=0.1,
#     keyheight=0.1,
#     default.unit="inch")
#   )
# pdf(file = "interp_beta1.pdf", family = "Times", height = 2.3, width = 3.7)
# print(pl)
# dev.off()
```

## Interpretation of $\beta_1$ **with** $X_2$ in model

![](./beta1_interp_1.pdf)

## Interpretation of $\beta_1$ **with** $X_2$ in model

![](./beta1_interp_2.pdf)

## Interpretation of $\beta_1$ **with** $X_2$ in model

![](./beta1_interp_3.pdf)

## Interpretation of $\beta_1$ **with** $X_2$ in model

![](./beta1_interp_4.pdf)

## Interpretation of $\beta_1$ **with** $X_2$ in model

![](./beta1_interp_5.pdf)

    
# Fitting Multiple Linear Regression in R

## How do we estimate the regression coefficients?

- Want to fit: $\mu(brain|gestation, body) = \beta_0 + \beta_1 gestation + \beta_2 body$

- $\beta_0$, $\beta_1$, $\beta_2$ are parameters (we don't know them)

- We can **estimate** them by minimizing the sum of the square residuals.

- Residuals: $Y_i - (\beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i})$

- The resulting estimates are the OLS (ordinary least squares) estimates: $\hat{\beta}_0$, $\hat{\beta}_1$, $\hat{\beta}_2$

## Syntax

- Want to fit: $\mu(brain|gestation, body) = \beta_0 + \beta_1 gestation + \beta_2 body$

- Use `lm()` and **always save the output**.

```{r, echo = TRUE}
lmout <- lm(Brain ~ Gestation + Body, data = case0902)
lmout
```

## Table

```{r}
lmout
```

- Interpreting output:

\begin{center}
\begin{tabular}{lll}
(Intercept) & Gestation & Body \\
$\hat{\beta}_0$ & $\hat{\beta}_1$ & $\hat{\beta}_2$\\
\end{tabular}
\end{center}

## Another Example

- Want to fit: $\mu(brain|gestation, body) = \beta_0 + \beta_1 gestation + \beta_2 body + \beta_3 litter$

```{r, echo = TRUE}
lmout <- lm(Brain ~ Gestation + Body + Litter, 
            data = case0902)
lmout
```

## Table

```{r}
lmout
```

- Interpreting output:

\begin{center}
\begin{tabular}{llll}
(Intercept) & Gestation & Body & Litter \\
$\hat{\beta}_0$ & $\hat{\beta}_1$ & $\hat{\beta}_2$ & $\hat{\beta}_3$\\
\end{tabular}
\end{center}

## Specific Language we Use

- We estimate that species with 1 day longer gestation time tend to have a brain weight 1.8 grams heavier, after adjusting for body weight and litter size.

- We estimate that species with an average body weight 1 kg heavier tend have a brain weight 0.99 grams heavier, after adjusting for gestation time and litter size.

- We estimate that species with a litter size of one offspring larger tend to have a brain weight 27.6 grams heavier, after adjusting for body weight and gestation time.

## Inference

- We are usually interested in testing if $\beta_i = 0$.

- We are usually interested in getting confidence intervals on the $\beta_i$'s.

- We can use the usual $t$-tools to get these.

## Inference in R

- $\mu(brain|gestation, body) = \beta_0 + \beta_1 gestation + \beta_2 body$

\footnotesize
```{r, echo = TRUE}
lmout <- lm(Brain ~ Gestation + Body, data = case0902)
summary(lmout)
```
\normalsize

## Inference in R

- $\mu(brain|gestation, body) = \beta_0 + \beta_1 gestation + \beta_2 body$

```{r}
sumout <- summary(lmout)
coef(sumout)
```

\begin{center}
\footnotesize
\begin{tabular}{lllll}
            & Estimate & Std. Error & t value & Pr(>|t|)\\
(Intercept) & $\hat{\beta}_0$ & $SE(\hat{\beta}_0)$ & $\hat{\beta}_0/SE(\hat{\beta}_0)$ & p-value for $H_0:\beta_0 = 0$ \\
Gestation   & $\hat{\beta}_1$ & $SE(\hat{\beta}_1)$ & $\hat{\beta}_1/SE(\hat{\beta}_1)$ & p-value for $H_0:\beta_1 = 0$ \\
Body        & $\hat{\beta}_2$ & $SE(\hat{\beta}_2)$ & $\hat{\beta}_2/SE(\hat{\beta}_2)$ & p-value for $H_0:\beta_2 = 0$ \\
\end{tabular}
\end{center}


## Confidence Intervals in R

```{r, echo = TRUE}
confint(lmout)
```

## Interpreting $p$-values


- The interpretations of significance depends on what other variables are in the model.

- $\mu(brain|gestation, body) = \beta_0 + \beta_1 gestation$

- $H_0: \beta_1 = 0$ vs $H_A: \beta_1 \neq 0$

- Model under Null: $\mu(brain|gestation, body) = \beta_0$

- Model under Alternative: $\mu(brain|gestation, body) = \beta_0 + \beta_1 gestation$

## Language we use

- If we reject $H_0$, then we say "we have strong evidence that gestation is related to brain weight."

- If we fail to reject $H_0$, then we say "we do not have strong evidence that gestation is related to brain weight."



## Interpreting $p$-values

- The interpretations of significance depends on what other variables are in the model.

- $\mu(brain|gestation, body) = \beta_0 + \beta_1 gestation + \beta_2 body$

- $H_0: \beta_1 = 0$ vs $H_A: \beta_1 \neq 0$

- Model under Null: $\mu(brain|gestation, body) = \beta_0 + \beta_2 body$

- Model under Alternative: $\mu(brain|gestation, body) = \beta_0 + \beta_1 gestation + \beta_2 body$

## Language we use

- If we reject $H_0$, then we say "we have strong evidence that gestation is related to brain weight after adjusting for body size."

- If we fail to reject $H_0$, then we say "we do not have strong evidence that gestation is related to brain weight after adjusting for body size."








