---
title: "Chapter 6 Worksheet"
author: "David Gerard"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: "blue"
bibliography: "../bib/614_bib.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = FALSE, fig.height = 3, fig.width = 4, fig.align = "center")
```


The `sardine` data frame from @clarke1936variations contains the number of vertebrae and the location of 12,858 sardines. The data are available on Blackboard (as "sardine.rdata") and you can load it into R using (you will probably have to change the path):
```{r, echo = TRUE}
load("../data/sardine.rdata")
head(sardine)
```

The `location` variable is coded by
\begin{center}
\begin{tabular}{l|l}
Code & Location \\
\hline
1 & Alaska \\
2 & British Columbia \\
3 & San Francisco \\
4 & Monterey \\
5 & San Pedro \\
6 & San Diego \\
\end{tabular}
\end{center}


1. Do you think the assumptions are satisfied enough that ANOVA is an appropriate model to use?

```{r}
library(ggplot2)
qplot(location, number, data = sardine, geom = "boxplot")
```

```{block}
Though the data are clearly not normal, there is a large enough sample size that we can appeal to the central limit theorem. The equal variance and independence assumptions are not fixed by a large enough sample size, but they seem to be valid.
```


2. Regardless of your answer to part 1, run a test to see if the means are the same.

```{r}
aout <- aov(number ~ location, data = sardine)
anova(aout)
```

```{block}
We have strong evidence that the means are different ($p$ = 0.00029).
```

3. If we wanted to compare all pairs of locations, how many tests would we be performing?

```{block}
This is 6 choose 2, $\binom{6}{2} = 6 \times 5 / 2 = 15$. You can calculate this in R using `choose(6, 2)`.
```


4. For what pairs of locations do we have evidence that the sardines have different numbers of vertebrae? (use Tukey here)

```{block}
Remember that we need to adjust for multiple comparisons. Using Tukey-adjusted $p$-values, it seems that we have the most evidence for a difference between locations 4 and 2, 4 and 6, and possibly 4 and 3.
```

```{r}
tout <- TukeyHSD(aout)
tout
```

5. If you would have used Holm-adjusted $p$-values instead, would the evidence for the pairwise differences in part 4 be as strong?

```{block}
The small $p$-values actually aren't that different. The large $p$-values seem to be a little larger, though.
```

```{r}
pairwise.t.test(sardine$number, sardine$location)
```

6. If the researchers would have suspected that vertebrae differ with colder climate *prior to looking at the data*, it might be reasonable to test if Alaskan and British Columbian sardines have different mean numbers of vertebrae. What model and hypotheses would you use to answer this question?

```{block}
- Model: $Y_{ij} = \mu_i + \epsilon_{ij}$
- $Y_{ij}$ is the number of vertebrae for the $j$th sardine in the $i$th location.
- $\mu_i$ is location $i$'s mean number of vertebrae.
- The $\epsilon_{ij}$'s are independent noise with common variance $\sigma^2$.

We would use
- $H_0: \frac{\mu_1 + \mu_2}{2} - \frac{\mu_3 + \mu_4 + \mu_5 + \mu_6}{4} = 0$
- $H_A: \frac{\mu_1 + \mu_2}{2} - \frac{\mu_3 + \mu_4 + \mu_5 + \mu_6}{4} \neq 0$
```

7. Test the hypotheses from part 6. State your conclusion.

```{r, message=FALSE}
## No evidence for a difference
library(car)
aout_2 <- aov(number ~ location - 1, data = sardine)
coef(aout_2)
coef_vec <- c(1/2, 1/2, -1/4, -1/4, -1/4, -1/4)
linearHypothesis(aout_2, hypothesis.matrix = coef_vec)
```







# References
