---
title: "Testing for Linear Combinations in Spock Example"
author: "David Gerard"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: "metropolis"
    slide_level: 2
urlcolor: "blue"
---

```{r setup, include=FALSE}
ggplot2::theme_set(
  ggplot2::theme_bw()
  )
knitr::opts_chunk$set(echo = TRUE,
                      fig.height = 2.5, 
                      fig.width = 4, 
                      fig.align = "center")
```

## Objectives

- Demonstrate how to interact with linear combinations of means in R.

- Analyze the Spock trial data in R.

## Load in Data
```{r}
library(Sleuth3)
library(ggplot2)
data("case0502")
```


## Spock EDA

```{r}
qplot(Judge, Percent, data = case0502)
```

## Spock EDA

```{r}
qplot(Judge, Percent, data = case0502, geom = "boxplot")
```

## Before Fitting
- Always make sure the grouping variable (the explanatory variable) is a "`factor`" with the `class()` function.
```{r}
class(case0502$Judge)
```
- Things will go wrong if this is any other type (even `logical` or `character` in the case of linear hypotheses)
- You can force a variable to be a `factor` with the `as.factor()` function:
```{r}
case0502$Judge <- as.factor(case0502$Judge)
```

## Fit the full model
- Use `aov()` function (for **A**nalysis **O**f **V**ariance) to fit the model that assumes $\mu_1$, $\mu_2$, \ldots, $\mu_I$ are all *different*.

- Always save this output.

- The **response** variable goes on the left of the tilde (`~`) and the **explanatory** variable goes to the right of the tilde.

## Fit the full model
```{r}
aout_alldiff <- aov(Percent ~ Judge, data = case0502)
aout_alldiff
```

## Wrong Parameterization

```{r}
coef(aout_alldiff)
```

- Returns estimates of $\mu$, $\delta_2$, $\delta_3$, $\delta_4$, $\delta_5$, $\delta_6$, $\delta_7$ in model where group $i$  has mean $\mu + \delta_i$.

## Do this to get correct parameterization
- **Subtracting 1** is notation for removing an "intercept column" and gives you the parameterization you expect.
\footnotesize
```{r}
aout_alldiff <- aov(Percent ~ Judge - 1, data = case0502)
coef(aout_alldiff)
```
\normalsize
- Returns estimates of $\mu_1$, $\mu_2$, $\mu_3$, $\mu_4$, $\mu_5$, $\mu_6$, $\mu_7$ in model where group $i$ has mean $\mu_i$. 
- **Never do this unless you are testing for linear combinations**
- The above code will **ruin** an anova table (when you call `anova()` or `summary()`).



## Linear Combinations

- The book says "A computer can produce the averages and the pooled estimate of variability, but hand calculations are usually required from there."
- But they don't know that there is *almost always* an R package that can do what you want (not true for any other statistical software).
- We will use the `linearHypothesis()` function from the `car` packages to run a test for a general linear combination of means.
```{r, eval=FALSE}
install.packages("car")
```
```{r, message=FALSE}
library(car)
```

## Look at order of factors

- Use the `levels()` command to see what the order of the factors is.

```{r}
levels(case0502$Judge)
```

## Set up coefficient vector
 
- The order is `("A", "B", "C", "D", "E", "F", "Spock's")`
- The following coefficient vector will test against
$$
H_0:~\text{Spock's} - \frac{1}{6}A - \frac{1}{6}B - \frac{1}{6}C - \frac{1}{6}D - \frac{1}{6}E - \frac{1}{6}F = 0
$$
```{r}
combo_vec1 <- c(-1/6, -1/6, -1/6, -1/6, -1/6, -1/6, 1)
```
- The following will test against
$$
H_0:~\frac{1}{3}A + \frac{1}{3}B + \frac{1}{3}C - \frac{1}{3}D - \frac{1}{3}E - \frac{1}{3}F = 0
$$
```{r}
combo_vec2 <- c(1/3, 1/3, 1/3, -1/3, -1/3, -1/3, 0)
```


## Run the hypothesis test
```{r}
lhout <- linearHypothesis(model = aout_alldiff, 
                          hypothesis.matrix = combo_vec1)
lhout
```

## Show equivalent to $t$-test
```{r, echo = FALSE, message=FALSE}
library(tidyverse)
case0502 %>%
  group_by(Judge) %>%
  summarize(mean = mean(Percent), sd = sd(Percent), n = n()) %>%
  ungroup() ->
  sumdat
gamma_hat <- sum(combo_vec1 * sumdat$mean)
sp <- sqrt(sum(aout_alldiff$residuals^2) / (nrow(case0502) - 7))
segamma_hat <- sp * sqrt(sum(combo_vec1^2 / sumdat$n))
tstat <- gamma_hat / segamma_hat
```

- $\hat{\gamma}$ turns out to be `r gamma_hat`
- $SE(\hat{\gamma})$ turns out to be `r segamma_hat`
- So the $t$-statistic is $\hat{\gamma} / SE(\hat{\gamma})$ = `r tstat`
- We compare this to a $t_{n - I} = t_{46 - 7} = t_{39}$
```{r}
2 * pt(-5.67, df = 39)
```
```{r}
lhout$`Pr(>F)`
```

## Correspondance

- The $t$-statistic is the square root of the $F$-statistic
- $t$-stat is -5.67
```{r}
sqrt(lhout$F)
```








